---
title: "Climate Effect on Pear Growth"
author: "Tamim Hussein"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Notes

-   Location is not included because the size_data only have 4 ecotron ids. All 4 ecotrons are in BE.

# Code Appenddix

```{r}
library(dplyr)
library(tidyverse)
library(car)
library(lme4)
library(tidyr)
library(lmerTest)
```

```{r}
size_data <- read.csv("size.data.csv")
soil_data <- read.csv("soil.data.csv")
outcome_data <- read.csv("outcome.data.csv")
```

```{r}
dim(size_data)
dim(outcome_data)
dim(soil_data)
```

## EDA

```{r}
colSums(is.na(size_data))
```

```{r}
size_data    <- na.omit(size_data)
soil_data    <- na.omit(soil_data)
outcome_data <- na.omit(outcome_data)
```

```{r}
size_data$species <- as.factor(size_data$species)
size_data$climate <- as.factor(size_data$climate)
size_data$ecotr_id<- as.factor(size_data$ecotr_id)
size_data$pear_id <- as.factor(size_data$pear_id)
size_data$tree_id <- as.factor(size_data$tree_id)
```

```{r}
table(size_data$ecotr_id, size_data$climate)
cat("\n\nNumber of unique species:", length(unique(na.omit(size_data$species))))
cat("\nUnique species:", unique(na.omit(size_data$species)))
cat("\n\nNumber of unique climates:", length(unique(na.omit(size_data$climate))))
cat("\nUnique climates:", unique(na.omit(size_data$climate)))
```

## Wide to Long Format

```{r}
# This transforms the data from wide format to long in terms of weeks
# while also changing weeks to numeric
long_df <- size_data %>%
  pivot_longer(
    cols = starts_with("week_"),
    names_to = "week_num",
    names_prefix = "week_",
    names_transform = list(week_num = ~ as.numeric(.) - 5),
    values_to = "size"
  )
```

```{r}
colnames(long_df)
```

```{r}
unique(long_df$week_num)
```

```{r}
head(long_df)
```

```{r}
ggplot(long_df,
       aes(x = week_num,
           color = interaction(species, climate))) +
  
  geom_line(aes(y = size,
                group = pear_id),
            alpha = 0.15,
            linewidth = 0.4) +
  
  stat_summary(aes(y = size,
                   group = interaction(species, climate)),
               fun = mean,
               geom = "line",
               linewidth = 1.4) +
  
  theme_minimal() +
  labs(
    x = "Week",
    y = "Predicted pear size",
    color = "Species + climate"
  )
```

## Linear Mixed Model

### Model 1

```{r}
# Linear + random intercept
model1 <- lmer(
  size ~ climate * species * week_num +
  (1 | pear_id),
  data = long_df
)
```

```{r}
summary(model1)
```

### Model 2

```{r}
# Linear + Random Intercept and Random slope
model2 <- lmer(
  size ~ climate * week_num * species +
  (1 + week_num| pear_id),
  data = long_df
)
```

```{r}
summary(model2)
```

### Model 3

```{r}
# Quadratic + random intercept
model3 <- lmer(size ~ climate * (week_num + I(week_num^2)) * species +
                 (1 | pear_id),
               data = long_df)
```

```{r}
summary(model3)
```

### Model 4

```{r}
# Quadratic with random intercept and random slope
model4 <- lmer(size ~ climate * (week_num + I(week_num^2)) * species +
                 (1 + week_num | pear_id),
               data = long_df)
```

```{r}
summary(model4)
```

### Model 5

```{r}
model4 <- lmer(size ~ climate * (week_num + I(week_num^2)) + species * week_num 
               + species * I(week_num^2)
               +(1 + week_num | pear_id),
               data = long_df)
```

```{r}
summary(model4)
```

### Model 6

```{r}
# Centering week to deal with week week^2 correlation
# This fixes quadratic mixed-model convergence issue
long_df$week_c <- long_df$week_num - mean(long_df$week_num)
```

```{r}
model5 <- lmer(
  size ~ climate * week_c +
    species * (week_c + I(week_c^2)) +
    I(week_c^2) +
    (1 + week_c | pear_id),
  data = long_df
)
```



```{r}
summary(model5)
```

```{r}
AIC(
  model1,
  model2,
  model3,
  model4,
  model5
)
```

```{r}
unique(long_df$week_c)
```


```{r}
names(summary(model5))
```

```{r}
summary(model1)$call
summary(model2)$call
summary(model3)$call
summary(model4)$call
summary(model4)$call
summary(model5)$call
```

## Model 6 Assumptions and Visualization

```{r}
library(ggplot2)

long_df$pred_individual <- predict(model5)

ggplot(long_df,
       aes(x = week_num,
           color = species)) +

  # individual pear predicted size
  geom_line(aes(y = pred_individual,
                group = pear_id),
            alpha = 0.15,
            linewidth = 0.4) +

  # average predicted size per species
  stat_summary(aes(y = pred_individual,
                   group = species),
               fun = mean,
               geom = "line",
               linewidth = 1.4,
               alpha = 1) +

  facet_wrap(~ climate) +
  theme_minimal() +
  labs(
    x = "Week",
    y = "Predicted pear size",
    color = "Species"
  )
```

```{r}
isSingular(model5)
```

```{r}
vc <- as.data.frame(VarCorr(model5))

pear_var <- vc$vcov[1]
resid_var <- vc$vcov[nrow(vc)]

ICC <- pear_var / (pear_var + resid_var)

ICC
```

```{r}
BIC(
  model1,
  model2,
  model3,
  model4,
  model5
)
```

```{r}
plot(model5)
qqnorm(resid(model5)); qqline(resid(model5))
```

```{r}
# Compare the estimates with model4 using another optimizer. Really powerful test!
# I am not sure if it is included in the course notes. Expect a question on it.
model_check <- update(
  model5,
  control = lmerControl(optimizer = "bobyqa")
)
```

```{r}
summary(model_check)
```

-   Model convergence was assessed using optimizer convergence diagnostics and singularity checks. No convergence issues or singular fits were detected.

-   The model appears well converged, non-singular, reasonably homoscedastic, approximately normal, well fitted, properly accounting for dependence

-   The estimated intraclass correlation coefficient (ICC) was approximately 0.47, indicating substantial within-pear dependence and supporting the use of a mixed-effects modeling framework.

-   BIC still strongly prefers model5, then the extra terms removed in earlier models were likely unnecessary

## Conclusions:

**The final model suggests:**

-   Pear size changes significantly over time.

-   Growth is nonlinear: growth increases over time but gradually slows down.

-   Pears exhibit substantial individual variability in both baseline size and growth trajectories, justifying the use of random intercepts and random slopes.

-   Climate scenarios affect growth trajectories mainly through differences in linear growth rates rather than curvature.

    -   Scenario 2 shows significantly faster growth than the reference climate.

    -   Scenario 4 also shows moderately increased growth.

    -   Scenario 3 behaves similarly to the reference climate.

-   Species differences are substantial.

    -   Doyenne pears are smaller on average than Conference pears.

    -   Doyenne pears grow more slowly over time.

    -   Doyenne pears also show stronger growth deceleration.

-   Diagnostic analyses indicated:

    -   acceptable homoscedasticity

    -   approximately normal residuals

    -   no major convergence or singularity issues

    -   good overall model fit

## Model 6 Extended Analysis

```{r}
library(emmeans)
```

```{r}
# Apply Scenario 2 vs Scenarios 1,3, and 4
emm <- emmeans(model5, ~ climate)

contrast(
  emm,
  list("Climate 2 vs average others" = c(-1/3, 1, -1/3, -1/3))
)
```

Scenario 2 has an estimated mean that is 0.621 units higher than the average of the estimated means of the other 3 scenarios.

```{r}
# Baseline Intercept per Pear on tree
eb <- ranef(model5)$pear_id
eb

plot(eb)
```

```{r}
library(geepack)
```

```{r}
# Generalized Estimating Equations (GEE) model
# with "exchangeable" correlation structure
gee1 <- geeglm(
  size ~ climate * week_c * species,
  id = pear_id,
  corstr = "exchangeable",
  data = long_df
)

gee1
```

GEE model gave similar estimates to the LMM, which is expected given the gaussian nature of the outcome variable.
